Rapid and Accurate Global PTM Discovery (G-PTM-D) Using Post-Acquisition Spectral Calibration and Defined Mass Windows

### 2017-06-20 paper.tex
yeah
Yahhhh

Many peptide residue modification types are known, and databases containing detailed information about such modifications are readily available\cite{Creasy_2004}.
Information about a modification often includes the chemical or isotopic composition, mass, specificity to certain residues, and possible restriction of placement to peptide or protein termini.
Despite the extensive cataloging of the different possible protein modifications, their comprehensive discovery and identifications in complex biological samples has continued to pose a difficult challenge for proteomics\cite{Olsen_2013}.

Numerous procedures for identification and localization of PTMs from "bottom-up" tandem mass spectrometric datasets exist, and those concentrate on easily studied modifications (e.g. phosphorylations) in well studied systems.
Global discovery tools, such as MODa\cite{Na_2011} are available.
G-PTM-D\cite{Li_2016}, is one recently described bioinformatics tool for the discovery and localization of new PTMs in unenriched samples that enables high confidence identification and discovery of a wide variety of different PTMs at the same time.
The G-PTM-D workflow consists of three stages: 1) A wide-mass database search\cite{Chick_2015, Na_2011} that provides spectral matches to unmodified peptides along with the mass differences between the identified peptides and the measured parent peptide masses (hypothesized to differ in mass due to the presence of a PTM).
2) For those peptides for which the mass difference corresponds to the mass of a known PTM, a database augmentation step adds plausible localized PTMs to the proteins in the search database.
3) A final standard narrow-mass search of the augmented database to identify both modified and unmodified peptides subject to the standard FDR threshold (e.g. 1\% FDR).

As described, G-PTM-D has a few significant limitations.
The initial wide-mass search procedure is slow, taking hours for modest size datasets.
Due to this, high mass modifications, such as glycosolations are out of reach, since widening the search window from the suggested $\pm 200$ Daltons increases the search time even more.
Furthermore, in G-PTM-D, modification discovery is limited to PTMs in the Uniprot database. 
Finally, PTMs that are similar in mass, such as phosphorylation and sulfonation are problematic since they are virtually indistinguishable in unprocessed spectra files.
These three limitations motivated the work described here.

Spectral calibration is essential for distinguishing modifications with similar mass, so we recommend running the G-PTM-D procedure only on calibrated spectra files.
We present an enhancement to G-PTM-D, that addresses the long search times of the wide-mass search, and the limitations of the standard final narrow-mass search.
MSFragger, a method recently described in\cite{Kong_2017}, significantly decreases the runtime of wide-mass searches, but lacks specificity.
Thererfore, we replace the wide-mass search in the first stage by a specialized search that only allows specific mass differences which correspond to pre-selected modification masses.
This change improves the specificity and significantly decreases search time.
The final narrow-mass search is also replaced by a specialized missed-monoisotope identifying search, further increasing the number of confidently identified peptides.

@@

\begin{figure}
  \includegraphics{diagram.eps}
  \caption{Workflows}
  \label{fgr:diagram}
\end{figure}

\section{Experimental Procedures}
We developed a modified version of the Morpheus software for bottom-up spectral database searching\cite{Wenger_2013} that we call MetaMorpheus\footnote{Free and Open Source software available at \url{https://github.com/smith-chem-wisc/MetaMorpheus}}, which integrates the database search procedure with spectral calibration and the G-PTM-D workflow.
Two multi-fraction mammalian datasets (\textit{Mouse} and \textit{Jurkat}) with deep proteome coverage were used to evaluate performance. These tryptically digested cell lysates correspond to 18 and 27 spectra files (fractions) respectively, and are described in detail in~\cite{Shortreed_2015, Cesnik_2016}.
Uniprot XML protein databases acquired on Feb 27, 2017 containing only reviewed human/mouse proteins were employed.

A 10ppm precursor mass tolerance was used for the initial calibration step, and then reduced to 5ppm for subsequent notch searches.
For the database augmentation step, we allowed 120 modifications including PTMs, adducts, and chemical modifications.
Final notch search used notches centered at \{0, 1.0029, 2.0052, 3.0077\} in order to capture missed identifications due to missed monoisotopic peaks.

@@

A critical parameter for peptide identification is mass accuracy\cite{Scherl_2008}.
Higher mass accuracy provides increased specificity and thus confidence in peptide identifications, decreasing the false discovery rate.
Instrument noise, systemic drift and miscalibration limit the mass accuracy in acquired spectra.
Multiple calibration strategies to improve mass accuracy have been devised, and fall into three general categories:  External calibration prior to the MS experiment (e.g. standard instrument calibration protocols); internal calibration during the MS experiment (e.g. real-time calibration using a lock mass standard\cite{Olsen_2005}); and subsequent to the MS experiment (post-acquisition spectral calibration).
We use a post-acquisition calibration procedure, that builds upon the software lock mass concept\cite{Cox_2011} recently reported by the Mann group.
In our strategy, the m/z differences between expected and observed peaks in the peptide tandem ms spectra are compiled, and then used to recalibrate the spectra.
%The increased mass accuracy of the calibrated spectra leads directly to improved identifications of both modified and unmodified peptides, as well as to increased confidence for PTM localization.

@@

\begin{figure}
  \includegraphics{fig2-calibrationQualityPhospho.eps}
  \caption{Histograms demonstrating the effect of calibration on PSM mass errors}
  \label{fgr:figure1}
\end{figure}
\newpage

@@

>>>>

Many peptide residue modification types are known, and databases containing detailed information about such modifications are readily available\cite{Creasy_2004}.
Information about a modification often includes the chemical or isotopic composition, mass, specificity to certain residues, and possible restriction of placement to peptide or protein termini.
\textbf{Protein databases such as UniProt may include localization information such as the observation of a modification on a specific amino acid in a protein, but this information is incomplete.}
Despite the extensive cataloging of the different possible protein modifications, their comprehensive discovery and identifications in complex biological samples has continued to pose a challenge for proteomics\cite{Olsen_2013}.

Numerous procedures for identification and localization of PTMs from "bottom-up" tandem mass spectrometric datasets exist, and those concentrate on common modifications (e.g. phosphorylations) in well studied systems.
Global discovery tools, such as MODa\cite{Na_2011} are also available, \textbf{but those have some drawbacks}.
G-PTM-D\cite{Li_2016}, is one recently described bioinformatics tool for high confidence identification and localization of a wide variety of different PTMs at the same time in unenriched samples.
The G-PTM-D workflow consists of three stages: 1) A wide-mass database search\cite{Chick_2015, Na_2011} that provides mass differences between the identified peptides and the measured parent peptide masses (hypothesized to differ in mass due to the presence of a PTM)
2) For those peptides for which the mass difference corresponds to the mass of a known PTM, a database augmentation step adds plausible localized PTMs to the proteins in the search database
3) A final standard narrow-mass search of the augmented database to identify both modified and unmodified peptides subject to the standard FDR threshold (e.g. 1\% FDR).

As described, G-PTM-D has a few significant limitations.
The initial wide-mass search procedure is slow, taking hours for modest size datasets.
Due to this, high mass modifications, such as glycosolations are out of reach, since widening the search window from the suggested $\pm 200$ Daltons increases the search time even more.
The final narrow-mass search is also problematic, since it does not account for mass errors due to missing the monoisotopic peak.
Furthermore, in G-PTM-D, modification discovery is limited to PTMs in the Uniprot database.
This is a significant limitation since that database does not include modifications arising from sample handling, and ignoring those significantly degrades the overall final identification rates.
Finally, PTMs that are similar in mass, such as phosphorylation and sulfonation are problematic since they are virtually indistinguishable in unprocessed spectra files.
These limitations motivated the work described here.

We enhance the G-PTM-D procedure in the following ways.
Spectral calibration is essential for distinguishing modifications with similar mass, so we recommend running the G-PTM-D procedure only on calibrated spectra files.
To address the long search times of the wide-mass search, we replace the wide-mass search in the first stage by a specialized search that only allows specific mass differences which correspond to pre-selected modification masses.
This change improves the specificity and significantly decreases search time.
The final narrow-mass search is also replaced by a specialized missed-monoisotope identifying search, further increasing the number of confidently identified peptides.
Finally, we expand the limited Uniprot PTM list with a customizeable expanded list of modifications that includes chemical adducts, glycosylations, lipids, and other important modifications.
This change not only expands the list of identified and localized modifications, but also indirectly improves the confident PSM count.

@@

\begin{figure}[H]
  \includegraphics{diagram.eps}
  \caption{Workflows}
  \label{fgr:diagram}
\end{figure}

\section{Experimental Procedures}
We developed a modified version of the Morpheus software for bottom-up spectral database searching\cite{Wenger_2013} that we call MetaMorpheus\footnote{Free and Open Source software available at \url{https://github.com/smith-chem-wisc/MetaMorpheus}}, which integrates the database search procedure with spectral calibration and the G-PTM-D workflow.
Two multi-fraction mammalian datasets (\textit{Jurkat} and \textit{Mouse}) with deep proteome coverage were used to evaluate performance.
These tryptically digested cell lysates correspond to 28 fractions from human Jurkat cells, and 18 fractions from Mouse islets, and are described in detail in~\cite{Shortreed_2015, Cesnik_2016}.

Uniprot XML protein databases acquired on Feb 27, 2017 containing only reviewed human/mouse proteins were employed.

A 10ppm precursor mass tolerance is used for the initial calibration step, and then reduced to 5ppm for subsequent notch searches.
For the database augmentation step, we allowed 120 modifications including PTMs, adducts, and chemical modifications.
Final notch search used notches centered at \{0, 1.0029, 2.0052, 3.0077\} in order to capture missed identifications due to missed monoisotopic peaks.

@@

A critical parameter for peptide identification is mass accuracy\cite{Scherl_2008}.
Higher mass accuracy provides increased specificity and thus confidence in peptide identifications, decreasing the false discovery rate.
Instrument noise, systemic drift and miscalibration limit the mass accuracy in acquired spectra.
Multiple calibration strategies to improve mass accuracy have been devised, and fall into three general categories:  External calibration prior to the MS experiment (e.g. standard instrument calibration protocols); internal calibration during the MS experiment (e.g. real-time calibration using a lock mass standard\cite{Olsen_2005}); and subsequent to the MS experiment (post-acquisition spectral calibration).
We use a post-acquisition calibration procedure that builds upon the software lock mass concept\cite{Cox_2011} recently reported by the Mann group.
In our strategy, the m/z differences between expected and observed peaks in the peptide tandem ms spectra are compiled, and then used to recalibrate the spectra.
%The increased mass accuracy of the calibrated spectra leads directly to improved identifications of both modified and unmodified peptides, as well as to increased confidence for PTM localization.

@@

\begin{figure}
  \includegraphics{fig2-calibrationQuality.eps}
  \caption{Histograms demonstrating the effect of calibration on PSM mass errors}
  \label{fgr:figure1}
\end{figure}
\newpage

@@

MSFragger, a method recently described in\cite{Kong_2017}, significantly decreases the runtime of wide-mass searches, but lacks specificity.
